Implementing Equational Constraints\\ in a Functional Language

### 2011-07-05 papers/wlp2011/paper.tex

@@

\usepackage{listings}
\lstset{aboveskip=1.0ex,
        belowskip=1.0ex,
        showstringspaces=false, % no special string space
        mathescape=true,
        flexiblecolumns=false,
        basewidth=0.52em,
        basicstyle=\small\ttfamily}

@@

In this paper we consider KiCS2 \cite{BrasselFischerHanusReck11},
a new system that compiles functional logic programs of the
source language Curry \cite{Hanus06Curry}
into purely functional Haskell programs.
We have shown in \cite{BrasselFischerHanusReck11}
that this implementation can compete with or outperform
other existing implementations of Curry.
KiCS2 is based on the idea to represent
the search space, i.e., all non-deterministic results of a computation,
as a data structure that can be traversed by operations
implementing various strategies.
Furthermore, logic variables are replaced by generators,
i.e., operations that non-deterministically evaluate to
all possible ground values of the type of the logic variable.
It has been shown \cite{AntoyHanus06ICLP}
that computing with logic variables by narrowing \cite{Reddy85,Slagle74}
or computing with generators by rewriting
are equivalent, i.e., compute the same results.
Although this implementation technique is correct \cite{Brassel11Thesis},
the generation of all values for logic variables
might be inefficient for applications that
exploit constraints on partial values.
For instance, in Prolog the equality constraint \ccode{X=c(a)}
is solved by instantiating the variable \code{X} to \code{c(a)},
but the equality constraint \ccode{X=Y} is solved by binding \code{X}
to \code{Y} without enumerating any values for \code{X} or \code{Y}.
Therefore, we propose in this paper new techniques
to implement equational constraints in the framework of KiCS2
(note that, in contrast to Prolog, unification is performed
modulo function evaluation).
Furthermore, we also show how functional patterns \cite{AntoyHanus05LOPSTR},
i.e., patterns containing evaluable operations providing
for more powerful pattern matching than in logic or functional languages,
can be implemented in this framework.
We show that both extensions lead to efficiency improvements
without sacrificing the efficiency of the kernel implementation.

@@

  xorSelf x = xor x x
\end{curry}
%
and the expression \ccode{xorSelf aBool}.
If we interpret this program as a term rewriting system,
we could have the reduction
\begin{haskell}
  xorSelf aBool  $\to~$  xor aBool aBool     $\to~$  xor True aBool
                 $\to~$  xor True False      $\to~$  True
\end{haskell}
leading to the unintended result \code{True}.
Note that this result cannot be obtained if we use a strict strategy
where arguments are evaluated prior to the function calls.
In order to avoid dependencies on the evaluation strategies
and exclude such unintended results,
Gonz\'alez-Moreno et al.\ \cite{GonzalezEtAl99} proposed
the rewriting logic CRWL as a logical
(execution- and strategy-independent) foundation for declarative
programming with non-strict and non-deterministic operations.  This
logic specifies the \emph{call-time choice} semantics \cite{Hussmann92}
\label{ctc-semantics}
where values of the arguments of an operation are determined before the
operation is evaluated. Note that this does not necessarily require
an eager evaluation of arguments.
Actually, \cite{AlbertHanusHuchOliverVidal05,LopezRodriguezSanchez07}
define lazy evaluation strategies for functional logic programs
with call-time choice semantics where actual arguments passed to
operations are shared. Hence, we can evaluate the expression above
lazily, provided that all occurrences of \code{aBool}
are shared so that all of them reduce either to \code{True} or to \code{False}
consistently. The requirements of this call-time choice semantics are the
reason why it is not simply possible to use list comprehensions
or non-determinism monads for a straightforward implementation
of functional logic programs in Haskell \cite{FischerKiselyovShan09}.

@@

In this section we sketch the translation of Curry programs
into Haskell programs as performed by KiCS2,
since this is necessary to understand the extensions
described in the subsequent sections.
More details about this translation scheme can be found in
\cite{BrasselFischer08IFL,BrasselFischerHanusReck11}.

As mentioned in the introduction, the KiCS2 implementation
is based on the explicit representation of non-deterministic results
in a data structure.
This is achieved by extending each data type of the source program
by constructors to represent a choice between two values
and a failure, respectively.
For instance, the data type for Boolean values defined in a Curry program by
\begin{curry}
  data Bool = False | True
\end{curry}
is translated into the Haskell data type\footnote{Actually,
our compiler performs some renamings to avoid conflicts with
predefined Haskell entities and introduces type classes
to resolve overloaded symbols like \code{Choice} and \code{Fail}.}
\begin{haskell}
  data Bool = False | True | Choice ID Bool Bool | Fail
\end{haskell}
The first argument of each \code{Choice} constructor of type \code{ID}
is used to implement the call-time choice semantics
discussed in Section~\ref{sec:Curry}.
Since the evaluation of \code{xorSelf aBool} duplicates
the argument operation \code{aBool}, we have to ensure
that both duplicates, which later evaluate to a non-deterministic
choice between two values, yield either \code{True} or \code{False}.
This is obtained by assigning a unique identifier (of type \code{ID})
to each \code{Choice}. The difficulty is to get a unique identifier
when needed, i.e., when some operation evaluates to a \code{Choice}.
Since we want to compile into \emph{purely} functional programs
(in order to enable powerful program optimizations),
we cannot use unsafe features with side effects to generate
such identifiers.
Hence, we pass a (conceptually infinite) set of identifiers,
also called \emph{identifier supply},
to each operation so that a \code{Choice} can pick an identifier from
this set.
For this purpose, we assume a type \code{IDSupply},
representing an infinite set of identifiers,
with operations
\begin{haskell}
  initSupply  :: IO IDSupply
  thisID      :: IDSupply -> ID
  leftSupply  :: IDSupply -> IDSupply
  rightSupply :: IDSupply -> IDSupply
\end{haskell}
The operation \code{initSupply} creates such a set (at the beginning
of an execution),
the operation \code{thisID} takes some identifier from this set, and
\code{leftSupply} and \code{rightSupply} split this set
into two disjoint subsets without the identifier
obtained by \code{thisID}.
There are different implementations available \cite{AugustssonRittriSynek94}
(see below for a simple implementation) and our implementation
is parametric over concrete implementations of \code{IDSupply}.

@@

In general, to extract all values from a \code{Choice} structure,
we have to traverse it, compute all possible choices
but consider the choice identifiers to make consistent (left/right)
decisions.
The latter requirement can be implemented by storing
the decisions already made for some choices during the traversal.
For this purpose, we introduce the type
\begin{haskell}
  data Decision = NoDecision | ChooseLeft | ChooseRight
\end{haskell}
where \code{NoDecision} represents the fact that the value of a choice
is yet undecided.
Furthermore, we assume operations to lookup the current decision
for a given identifier or change its decision (depending on the implementation
of \code{IDSupply}, KiCS2 supports various implementations
based on memory cells or finite maps):
\begin{haskell}
  lookupDecision :: ID -> IO Decision
  setDecision    :: ID -> Decision -> IO ()
\end{haskell}
%
Now we can print all values contained in a choice structure
in a depth-first manner by the following operation:\footnote{%
Note that this code has been simplified for readability
since the type system of Haskell does not
allow this direct definition.}
\label{sec:printValsDFS}
\begin{haskell}
  printValsDFS :: a -> IO ()
  printValsDFS Fail             = return ()
  printValsDFS (Choice i x1 x2) = lookupDecision i >>= follow
  printValsDFS v                = print v
   where
    follow ChooseLeft  = printValsDFS x1
    follow ChooseRight = printValsDFS x2
    follow NoDecision  = do newDecision ChooseLeft  x1
                            newDecision ChooseRight x2

    newDecision d x = do setDecision i d
                         printValsDFS x
                         setDecision i NoDecision
\end{haskell}
This operation ignores failures and prints values that are no \code{Choice}.
For a \code{Choice}, it checks whether a decision for
this identifier has already been made (note that the initial value
for all identifiers is \code{NoDecision}).
If a decision has been made for this choice, it follows this decision.
Otherwise, the left alternative is used and this decision is stored.
After printing
all the values w.r.t.\ this decision,
the decision is undone (like in backtracking)
and the right alternative is used and stored.

@@

To avoid an unnecessary growth of the search space represented by
\code{Choice} constructors, out compiler performs an optimization for
deterministic operations. If an operation does not call, 
neither directly nor indirectly through
other operations, the choice operation \ccode{?},\footnote{%
Note that before applying this compilation scheme, overlapping rules in
user-defined operations are transformed into calls of the choice operation
\ccode{?}.}
then it is not necessary to pass an identifier supply to the operation.
In consequence, only the matching rules are extended so that the generated code
is nearly identical to a corresponding functional program.
Actually, the benchmarks presented in \cite{BrasselFischerHanusReck11}
show that this implementation outperforms all other Curry implementations
for deterministic operations,
and, for non-deterministic operations, outperforms Prolog-based
implementations of Curry and can compete with MCC \cite{Lux99FLOPS},
a Curry implementation that compiles to C.

@@

Due to this constructive definition, \ccode{=:=}
can be considered as a binary function defined by the
following rules (here we show only the rules for the Boolean
and list types, where \code{Success} denotes the only constructor
of the type \code{Success} of constraints):
\begin{curry}
  True   =:= True    =  Success
  False  =:= False   =  Success

  []     =:= []      =  Success
  (x:xs) =:= (y:ys)  =  x =:= y & xs =:= ys

@@

\end{haskell}
%
Although this is a correct implementation of equational constraints,
it might lead to an unnecessarily large search space
when it is applied to generators representing logic variables.
For instance, consider the following generator for Boolean lists:
\begin{curry}
  aBoolList = [] ? (aBool : aBoolList)
\end{curry}
This is translated into Haskell as follows:
\begin{haskell}
  aBoolList :: IDSupply -> [Bool]
  aBoolList s = Choice (thisID s) [] (aBool (leftSupply s)
                                      : aBoolList (rightSupply s))
\end{haskell}
Now consider the equational constraint \ccode{x =:= [True]}.
If the logic variable \code{x} is replaced by \code{aBList},
the translated expression \ccode{aBoolList s =:= [True]}
creates a search space when evaluating its first argument,
although there is no search required since there is only one possible
binding for \code{x}.
Furthermore and even worse, there is an infinite search space
when unifying two logic variables. For instance,
the expression \ccode{xs =:= ys \& xs++ys =:= [True]}
results in an infinite search space when the logic variables
\code{xs} and \code{ys} are replaced by generators.

@@

To implement unification, we have to distinguish free variables
from ``standard choices'' (introduced by the operation \ccode{?})
in the target code. For this purpose, we refine the definition
of the type \code{ID} as follows:\footnote{For the sake
of simplicity, we consider in the following the
implementation of \code{IDSupply} as unbounded integers.}
\begin{haskell}
  data ID = ChoiceID Integer | FreeID Integer
\end{haskell}
The new constructor \code{FreeID} marks a choice corresponding to
a free variable, e.g., the generator for Boolean variables is
redefined as
\begin{haskell}
  aBool s = Choice (FreeID (thisID s)) True False
\end{haskell}
%
If an operation is applied to a free variable and requires its value,
the free variable is transformed into a standard choice.
For this purpose, we define a simple operation to perform
this transformation:
\begin{haskell}
  narrow :: ID -> ID
  narrow (FreeID i) = ChoiceID i
  narrow x          = x
\end{haskell}
Furthermore, we use this operation in narrowing steps,
i.e., in all rules operating on \code{Choice} constructors.
For instance, in the implementation of the operation \code{not}
we replace the rule
\begin{haskell}
  not (Choice i x1 x2) s = Choice i (not x1 s) (not x2 s)
\end{haskell}
by the rule
\begin{haskell}
  not (Choice i x1 x2) s = Choice (narrow i) (not x1 s) (not x2 s)
\end{haskell}
With this small change, the run-time system can distinguish
free variables (choices with an identifier of the form \code{(FreeID $\ldots$)})
from other values. This is quite useful at the top-level
to print values containing logic variables in a specific form
rather than enumerating all values, similarly to logic programming systems.
For instance, KiCS2 evaluates the application of \code{head}
to an unknown list as follows:
\begin{haskell}
  Prelude> head xs where xs free
  {xs = (_x2:_x3)} _x2
\end{haskell}
Thus, free variables are prefixed by \code{\us{}x}.

As mentioned above, the consideration of free variables
is relevant in equational constraints where \emph{binding constraints} are
generated. For this purpose, we introduce a type to represent
a binding constraint as a pair of a choice identifier
and a decision for this choice:
\begin{haskell}
  data Constraint = ID :=: Decision
\end{haskell}
Furthermore, we extend each data type by the possibility to
add such a constraint:
\begin{haskell}
  data Bool   = $\ldots$ | Guard Constraint Bool
  data List a = $\ldots$ | Guard Constraint (List a)
\end{haskell}
Thus, \code{(Guard $c$ $v$)} represents a \emph{constrained value},
i.e., the value $v$ is only valid if the constraint $c$ is
consistent with the decisions previously made during search.
The binding constraints are created by the
equational constraint operation \ccode{=:=}: if a free variable
should be bound to a constructor, we make the same
decision as it would be done in the successful branch
of the generator. In case of Boolean values,
this can be implemented by the following additional rules
for \ccode{=:=}:
\begin{haskell}
  Choice (FreeID i) _ _ =:= True   =  Guard (i :=: ChooseLeft ) Success
  Choice (FreeID i) _ _ =:= False  =  Guard (i :=: ChooseRight) Success
\end{haskell}
Hence, the binding of a variable to some known value
is implemented as a binding constraint for the choice identifier
for this variable. However, if we want to bind a variable
to another variable, we cannot store a concrete decision.
Instead, we store the information that the decisions for
both variables, when they are made to extract values,
must be identical. For this purpose, we extend the \code{Decision}
type to cover this information:
\begin{haskell}
  data Decision = $\ldots$ | BindTo ID
\end{haskell}
Furthermore, we add the rule that an equational constraint
between two variables yields a binding for these variables:
\begin{haskell}
  Choice (FreeID i) _ _ =:= Choice (FreeID j) _ _
    =  Guard (i :=: BindTo j) Success
\end{haskell}
%

The consistency of constraints is checked when values are extracted
from a choice structure, e.g., by the operation \code{printValsDFS}.
For this purpose, we extend the definition of the corresponding search
operations by calling a constraint solver for the constraints.
For instance, the definition of \code{printValsDFS} is extended by
a rule handling constrained values:
%
\begin{haskell}
  $\ldots$
  printValsDFS (Guard c x) = do consistent <- add c
                                if consistent then do printValsDFS x
                                                      remove c
                                              else return ()
  $\ldots$
\end{haskell}
The operation \code{add} checks the consistency of the constraint \code{c}
with the decisions made so far and stores the decision made by the constraint
if consistency is sustained. In this case, the constrained value is evaluated
before the contstraint is removed to allow backtracking.
Furthermore, the operations \code{lookupDecision} and \code{setDecision}
are extended to deal with bindings between two variables,
i.e., they follow variable chains in case of \code{BindTo} constructors.

@@

A well-known disadvantage of equational constraints
is the fact that \ccode{=:=} is interpreted as strict equality.
Thus, if one uses equational constraints to express
requirements on arguments, the resulting operations might be too strict.
For instance, the equational constraint in the condition
defining \code{last} (see Section~\ref{ex:last})
requires that \code{ys++[e]} as well as \code{xs}
must be reducible to identical values so that the input list \code{xs}
is completely evaluated.
Hence, if \code{failed} denotes an operation whose evaluation fails,
the evaluation of \code{last [failed,True]} has no result.
On the other hand, the evaluation of \code{last' [failed,True]} yields
the value \code{True}, i.e., the definition of \code{last'} is less strict.
This is due to the fact that a functional pattern like
\code{(xs++[e])} abbreviates all values to which it can be evaluated
(by narrowing), like \code{[e]}, \code{[x1,e]}, \code{[x1,x2,e]} etc.
Conceptually, the rule defining \code{last'}
abbreviates the following (infinite) set of rules:
\begin{curry}
  last' [e] = e
  last' [x1,e] = e
  last' [x1,x2,e] = e
  $\ldots$
\end{curry}
As a further example,
consider an operation that returns the first duplicated element in
a list.
Using equational constraint, we can define it as follows:
\begin{curry}
  fstDup xs | xs =:= ys++[e]++zs & elem e ys =:= True & nub ys =:= ys
            = e    where ys, zs, e free
\end{curry}
The first equational constraint is used to split the input list \code{xs}
into three sublists.
The last constraint ensures that the first sublist
\code{ys} does not contain duplicated elements
(the library operation \code{nub} removes all duplicates from a list)
and the second constraint ensures that the first element after \code{ys}
occurs in \code{ys}.
Although this implementation is concise, it cannot be applied
to infinite lists due to the strict interpretation of \ccode{=:=}.
This is not the case if we define this operation by a functional pattern:
\begin{curry}
  fstDup' (ys++[e]++zs) | elem e ys =:= True & nub ys =:= ys
                        = e
\end{curry}
%
Functional patterns are quite useful in situations
where one wants to select parts of some structure
at arbitrary positions. For instance,
functional patterns have been applied to support
the high-level programming of querying or transforming
XML data \cite{Hanus11ICLP}.

@@

In order to demonstrate the operational behavior of our implementation,
we sketch the evaluation of the lazy unification constraint
\code{xs++[e] =:<= [failed,True]} that occurs when the expression
\code{last' [failed,True]} is evaluated (here we omit failed branches
and some other details; note that logic variables are replaced
by generators):
\begin{haskell}
  aBoolList 2 ++ [aBool 3] =:<= [failed,True]
  $\leadsto$ [aBool 4,aBool 3] =:<= [failed,True]
  $\leadsto$ aBool 4 =:<= failed & aBool 3 =:<= True & [] =:<= []
  $\leadsto$ Guard (4 :=: LazyBind (lazyBind 4 failed))
           (Guard (3 :=: LazyBind (lazyBind 3 True)) Success)
\end{haskell}
If the binding of identifier \code{3} is later required,
\code{(lazyBind 3 True)} is evaluated to \code{[3 :=: ChooseLeft]}
which corresponds to the value \code{True} of the generator \code{aBool}.
Note that the variable with identifier \code{4} does not occur
anywhere else so that the binding \code{(lazyBind 4 failed)}
will never be evaluated, as intended.

>>>>

\addtolength{\textheight}{0.2cm}
\addtolength{\topmargin}{-0.1cm}
\addtolength{\textwidth}{0.2cm}
\addtolength{\oddsidemargin}{-0.1cm}
\addtolength{\evensidemargin}{-0.1cm}

@@

\usepackage{listings}
\lstset{aboveskip=0.5ex,
        belowskip=0.5ex,
        showstringspaces=false, % no special string space
        mathescape=true,
        flexiblecolumns=false,
        basewidth=0.52em,
        basicstyle=\small\ttfamily}

@@

In this paper we consider KiCS2 \cite{BrasselFischerHanusReck11},
a new system that compiles functional logic programs of the
source language Curry \cite{Hanus06Curry}
into purely functional Haskell programs.
We have shown in \cite{BrasselFischerHanusReck11}
that this implementation can compete with or outperform
other existing implementations of Curry.
KiCS2 is based on the idea to represent
the search space, i.e., all non-deterministic results of a computation,
as a data structure that can be traversed by operations
implementing various strategies.
Furthermore, logic variables are replaced by generators,
i.e., operations that non-deterministically evaluate to
all possible ground values of the type of the logic variable.
It has been shown \cite{AntoyHanus06ICLP}
that computing with logic variables by narrowing \cite{Reddy85,Slagle74}
or computing with generators by rewriting
are equivalent, i.e., compute the same values.
Although this implementation technique is correct \cite{Brassel11Thesis},
the generation of all values for logic variables
might be inefficient for applications that
exploit constraints on partial values.
For instance, in Prolog the equality constraint \ccode{X=c(a)}
is solved by instantiating the variable \code{X} to \code{c(a)},
but the equality constraint \ccode{X=Y} is solved by binding \code{X}
to \code{Y} without enumerating any values for \code{X} or \code{Y}.
Therefore, we propose in this paper new techniques
to implement equational constraints in the framework of KiCS2
(note that, in contrast to Prolog, unification is performed
modulo function evaluation).
Furthermore, we also show how functional patterns \cite{AntoyHanus05LOPSTR},
i.e., patterns containing evaluable operations providing
for more powerful pattern matching than in logic or functional languages,
can be implemented in this framework.
We show that both extensions lead to efficiency improvements
without sacrificing the efficiency of the kernel implementation.

@@

  xorSelf x = xor x x
\end{curry}
%
and the expression \ccode{xorSelf aBool}.
If we interpret this program as a term rewriting system,
we could have the reduction
\begin{haskell}
  xorSelf aBool  $\to~$  xor aBool aBool     $\to~$  xor True aBool
                 $\to~$  xor True False      $\to~$  True
\end{haskell}
leading to the unintended result \code{True}.
Note that this result cannot be obtained if we use a strict strategy
where arguments are evaluated prior to the function calls.
In order to avoid dependencies on the evaluation strategies
and exclude such unintended results,
Gonz\'alez-Moreno et al.\ \cite{GonzalezEtAl99} proposed
the rewriting logic CRWL as a logical
(execution- and strategy-independent) foundation for declarative
programming with non-strict and non-deterministic operations.  This
logic specifies the \emph{call-time choice} semantics \cite{Hussmann92}
\label{ctc-semantics}
where values of the arguments of an operation are determined before the
operation is evaluated. This can be enforced in a lazy strategy
by sharing actual arguments.
For instance, the expression above can be lazily evaluated
provided that all occurrences of \code{aBool}
are shared so that all of them reduce either to \code{True} or to \code{False}
consistently.

@@

To understand the extensions described in the subsequent sections,
we sketch the translation of Curry programs
into Haskell programs as performed by KiCS2.
More details about this translation scheme can be found in
\cite{BrasselFischer08IFL,BrasselFischerHanusReck11}.

As mentioned in the introduction, the KiCS2 implementation
is based on the explicit representation of non-deterministic results
in a data structure.
This is achieved by extending each data type of the source program
by constructors to represent a choice between two values
and a failure, respectively.
For instance, the data type for Boolean values defined in a Curry program by
\begin{curry}
  data Bool = False | True
\end{curry}
is translated into the Haskell data type\footnote{Actually,
our compiler performs some renamings to avoid conflicts with
predefined Haskell entities and introduces type classes
to resolve overloaded symbols like \code{Choice} and \code{Fail}.}
\begin{haskell}
  data Bool = False | True | Choice ID Bool Bool | Fail
\end{haskell}
The first argument of type \code{ID} of each \code{Choice} constructor
is used to implement the call-time choice semantics
discussed in Section~\ref{sec:Curry}.
Since the evaluation of \code{xorSelf aBool} duplicates
the argument operation \code{aBool}, we have to ensure
that both duplicates, which later evaluate to a non-deterministic
choice between two values, yield either \code{True} or \code{False}.
This is obtained by assigning a unique identifier (of type \code{ID})
to each \code{Choice}. The difficulty is to get a unique identifier
when needed, i.e., when some operation evaluates to a \code{Choice}.
Since we want to compile into \emph{purely} functional programs
(in order to enable powerful program optimizations),
we cannot use unsafe features with side effects to generate
such identifiers.
Hence, we pass a (conceptually infinite) set of identifiers,
also called \emph{identifier supply},
to each operation so that a \code{Choice} can pick an identifier from
this set.
For this purpose, we assume a type \code{IDSupply},
representing an infinite set of identifiers,
with operations
\begin{haskell}
  initSupply  :: IO IDSupply
  thisID      :: IDSupply -> ID
  leftSupply  :: IDSupply -> IDSupply
  rightSupply :: IDSupply -> IDSupply
\end{haskell}
The operation \code{initSupply} creates such a set (at the beginning
of an execution),
the operation \code{thisID} takes some identifier from this set, and
\code{leftSupply} and \code{rightSupply} split this set
into two disjoint subsets without the identifier
obtained by \code{thisID}.
There are different implementations available \cite{AugustssonRittriSynek94}
(see below for a simple implementation) and our implementation
is parametric over concrete implementations of \code{IDSupply}.

@@

In general, to extract all values from a \code{Choice} structure,
we have to traverse it, compute all possible choices
but consider the choice identifiers to make consistent (left/right)
decisions.
The latter requirement can be implemented by storing
the decisions already made for some choices during the traversal.
For this purpose, we introduce the type
\begin{haskell}
  data Decision = NoDecision | ChooseLeft | ChooseRight
\end{haskell}
where \code{NoDecision} represents the fact that the value of a choice
is yet undecided.
Furthermore, we assume operations to lookup the current decision
for a given identifier or change its decision (depending on the implementation
of \code{IDSupply}, KiCS2 supports various implementations
based on memory cells or finite maps):
\begin{haskell}
  lookupDecision :: ID -> IO Decision
  setDecision    :: ID -> Decision -> IO ()
\end{haskell}
%
Now we can print all values contained in a choice structure
in a depth-first manner by the following I/O operation:\footnote{%
Note that this code has been simplified for readability
since the type system of Haskell does not
allow this direct definition.}
\label{sec:printValsDFS}
\begin{haskell}
  printValsDFS :: a -> IO ()
  printValsDFS Fail             = return ()
  printValsDFS (Choice i x1 x2) = lookupDecision i >>= follow
  printValsDFS v                = print v
   where
    follow ChooseLeft  = printValsDFS x1
    follow ChooseRight = printValsDFS x2
    follow NoDecision  = do newDecision ChooseLeft  x1
                            newDecision ChooseRight x2

    newDecision d x = do setDecision i d
                         printValsDFS x
                         setDecision i NoDecision
\end{haskell}
This operation ignores failures and prints values that are not a \code{Choice}.
For a \code{Choice}, it checks whether a decision for
this identifier has already been made (note that the initial value
for all identifiers is \code{NoDecision}).
If a decision has been made for this choice, it follows this decision.
Otherwise, the left alternative is used and this decision is stored.
After printing
all the values w.r.t.\ this decision,
the decision is undone (like in backtracking)
and the right alternative is used and stored.

@@

To avoid an unnecessary growth of the search space represented by
\code{Choice} constructors, our compiler performs an optimization for
deterministic operations. If an operation does not call, 
neither directly nor indirectly through
other operations, the choice operation \ccode{?},\footnote{%
Note that before applying this compilation scheme, overlapping rules in
user-defined operations are transformed into calls of the choice operation
\ccode{?}.}
then it is not necessary to pass an identifier supply to the operation.
In consequence, only the matching rules are extended so that the generated code
is nearly identical to a corresponding functional program.
Actually, the benchmarks presented in \cite{BrasselFischerHanusReck11}
show that this implementation outperforms all other Curry implementations
for deterministic operations,
and, for non-deterministic operations, outperforms Prolog-based
implementations of Curry and can compete with MCC \cite{Lux99FLOPS},
a Curry implementation that compiles to C.

@@

Due to this constructive definition, \ccode{=:=}
can be considered as a binary function defined by the
following rules (here we show only the rules for the Boolean
and list types, where \code{Success} denotes the only constructor
of the type \code{Success} of constraints):
\begin{curry}
  True   =:= True    =  Success
  False  =:= False   =  Success
  []     =:= []      =  Success
  (x:xs) =:= (y:ys)  =  x =:= y & xs =:= ys

@@

\end{haskell}
%
Although this is a correct implementation of equational constraints,
it might lead to an unnecessarily large search space
when it is applied to generators representing logic variables.
For instance, consider the following generator for Boolean lists:
\begin{curry}
  aBoolList = [] ? (aBool : aBoolList)
\end{curry}
This is translated into Haskell as follows:
\begin{haskell}
  aBoolList :: IDSupply -> [Bool]
  aBoolList s = Choice (thisID s) [] (aBool (leftSupply s)
                                      : aBoolList (rightSupply s))
\end{haskell}
Now consider the equational constraint \ccode{x =:= [True]}.
If the logic variable \code{x} is replaced by \code{aBoolList},
the translated expression \ccode{aBoolList s =:= [True]}
creates a search space when evaluating its first argument,
although there is no search required since there is only one possible
binding for \code{x}.
Furthermore and even worse, there is an infinite search space
when unifying two logic variables. For instance,
the expression \ccode{xs =:= ys \& xs++ys =:= [True]}
results in an infinite search space when the logic variables
\code{xs} and \code{ys} are replaced by generators.

@@

To implement unification, we have to distinguish free variables
from ``standard choices'' (introduced by the operation \ccode{?})
in the target code. For this purpose, we refine the definition
of the type \code{ID} as follows:\footnote{For the sake
of simplicity, we consider in the following the
implementation of \code{IDSupply} as unbounded integers.}
\begin{haskell}
  data ID = ChoiceID Integer | FreeID Integer
\end{haskell}
The new constructor \code{FreeID} marks a choice corresponding to
a free variable, e.g., the generator for Boolean variables is
redefined as
\begin{haskell}
  aBool s = Choice (FreeID (thisID s)) True False
\end{haskell}
%
If an operation is applied to a free variable and requires its value,
the free variable is transformed into a standard choice.
For this purpose, we define a simple operation to perform
this transformation:
\begin{haskell}
  narrow :: ID -> ID
  narrow (FreeID i) = ChoiceID i
  narrow x          = x
\end{haskell}
Furthermore, we use this operation in narrowing steps,
i.e., in all rules operating on \code{Choice} constructors.
For instance, in the implementation of the operation \code{not}
we replace the rule
\begin{haskell}
  not (Choice i x1 x2) s = Choice i (not x1 s) (not x2 s)
\end{haskell}
by the rule
\begin{haskell}
  not (Choice i x1 x2) s = Choice (narrow i) (not x1 s) (not x2 s)
\end{haskell}
With this small change, the run-time system can distinguish
free variables (choices with an identifier of the form \code{(FreeID $\ldots$)})
from other values. This is quite useful at the top-level
to print values containing logic variables in a specific form
rather than enumerating all values, similarly to logic programming systems.
For instance, KiCS2 evaluates the application of \code{head}
to an unknown list as follows:
\begin{haskell}
  Prelude> head xs where xs free
  {xs = (_x2:_x3)} _x2
\end{haskell}
Here, free variables are marked by the prefix \code{\us{}x}.

As mentioned above, the consideration of free variables
is relevant in equational constraints where \emph{binding constraints} are
generated. For this purpose, we introduce a type to represent
a binding constraint as a pair of a choice identifier
and a decision for this choice:
\begin{haskell}
  data Constraint = ID :=: Decision
\end{haskell}
Furthermore, we extend each data type by the possibility to
add such a constraint:
\begin{haskell}
  data Bool   = $\ldots$ | Guard Constraint Bool
  data List a = $\ldots$ | Guard Constraint (List a)
\end{haskell}
Thus, \code{(Guard $c$ $v$)} represents a \emph{constrained value},
i.e., the value $v$ is only valid if the constraint $c$ is
consistent with the decisions previously made during search.
The binding constraints are created by the
equational constraint operation \ccode{=:=}: if a free variable
should be bound to a constructor, we make the same
decision as it would be done in the successful branch
of the generator. In case of Boolean values,
this can be implemented by the following additional rules
for \ccode{=:=}:
\begin{haskell}
  Choice (FreeID i) _ _ =:= True   =  Guard (i :=: ChooseLeft ) Success
  Choice (FreeID i) _ _ =:= False  =  Guard (i :=: ChooseRight) Success
\end{haskell}
Hence, the binding of a variable to some known value
is implemented as a binding constraint for the choice identifier
for this variable. However, if we want to bind a variable
to another variable, we cannot store a concrete decision.
Instead, we store the information that the decisions for
both variables, when they are made to extract values,
must be identical. For this purpose, we extend the \code{Decision}
type to cover this information:
\begin{haskell}
  data Decision = $\ldots$ | BindTo ID
\end{haskell}
Furthermore, we add the rule that an equational constraint
between two variables yields a binding for these variables:
\begin{haskell}
  Choice (FreeID i) _ _ =:= Choice (FreeID j) _ _
    =  Guard (i :=: BindTo j) Success
\end{haskell}
%
The consistency of constraints is checked when values are extracted
from a choice structure, e.g., by the operation \code{printValsDFS}.
For this purpose, we extend the definition of the corresponding search
operations by calling a constraint solver for the constraints.
For instance, the definition of \code{printValsDFS} is extended by
a rule handling constrained values:
%
\begin{haskell}
  $\ldots$
  printValsDFS (Guard c x) = do consistent <- add c
                                if consistent then do printValsDFS x
                                                      remove c
                                              else return ()
  $\ldots$
\end{haskell}
The operation \code{add} checks the consistency of the constraint \code{c}
with the decisions made so far and, in case of consistency,
stores the decision made by the constraint.
In this case, the constrained value is evaluated
before the constraint is removed to allow backtracking.
Furthermore, the operations \code{lookupDecision} and \code{setDecision}
are extended to deal with bindings between two variables,
i.e., they follow variable chains in case of \code{BindTo} constructors.

@@

A well-known disadvantage of equational constraints
is the fact that \ccode{=:=} is interpreted as strict equality.
Thus, if one uses equational constraints to express
requirements on arguments, the resulting operations might be too strict.
For instance, the equational constraint in the condition
defining \code{last} (see Section~\ref{ex:last})
requires that \code{ys++[e]} as well as \code{xs}
must be reducible to identical values so that the input list \code{xs}
is completely evaluated.
Hence, if \code{failed} denotes an operation whose evaluation fails,
the evaluation of \code{last [failed,True]} has no result.
On the other hand, the evaluation of \code{last' [failed,True]} yields
the value \code{True}, i.e., the definition of \code{last'} is less strict.
This is due to the fact that a functional pattern like
\code{(xs++[e])} abbreviates all values to which it can be evaluated
(by narrowing), like \code{[e]}, \code{[x1,e]}, \code{[x1,x2,e]} etc.
Conceptually, the rule defining \code{last'}
abbreviates the following (infinite) set of rules:
\begin{curry}
  last' [e] = e
  last' [x1,e] = e
  last' [x1,x2,e] = e
  $\ldots$
\end{curry}
As a further example,
consider an operation that returns the first duplicated element in
a list.
Using equational constraints, we can define it as follows:
\begin{curry}
  fstDup xs | xs =:= ys++[e]++zs & elem e ys =:= True & nub ys =:= ys
            = e    where ys, zs, e free
\end{curry}
The first equational constraint is used to split the input list \code{xs}
into three sublists.
The last constraint ensures that the first sublist
\code{ys} does not contain duplicated elements
(the library operation \code{nub} removes all duplicates from a list)
and the second constraint ensures that the first element after \code{ys}
occurs in \code{ys}.
Although this implementation is concise, it cannot be applied
to infinite lists due to the strict interpretation of \ccode{=:=}.
This is not the case if we define this operation by a functional pattern:
\begin{curry}
  fstDup' (ys++[e]++zs) | elem e ys =:= True & nub ys =:= ys
                        = e
\end{curry}
%
Functional patterns are quite useful in situations
where one wants to select parts of some structure
at arbitrary positions. For instance,
functional patterns have been applied to support
the high-level programming of querying or transforming
XML data \cite{Hanus11ICLP}.

@@

In order to demonstrate the operational behavior of our implementation,
we sketch the evaluation of the lazy unification constraint
\code{xs++[e] =:<= [failed,True]} that occurs when the expression
\code{last' [failed,True]} is evaluated (here we omit failed branches
and some other details; note that logic variables are replaced
by generators):
\begin{haskell}
  aBoolList 2 ++ [aBool 3] =:<= [failed,True]
  $\leadsto$ [aBool 4,aBool 3] =:<= [failed,True]
  $\leadsto$ aBool 4 =:<= failed & aBool 3 =:<= True & [] =:<= []
  $\leadsto$ Guard (4 :=: LazyBind (lazyBind 4 failed))
           (Guard (3 :=: LazyBind (lazyBind 3 True)) Success)
\end{haskell}
If the binding of identifier \code{3} is later required,
\code{(lazyBind 3 True)} is evaluated to \code{[3 :=: ChooseLeft]}
which corresponds to the value \code{True} of the generator \code{(aBool 3)}.
Note that the variable with identifier \code{4} does not occur
anywhere else so that the binding \code{(lazyBind 4 failed)}
will never be evaluated, as intended.
